An overview on temporal databases.\\ New approaches and proposals.

### 2011-11-09 trunk/book/sources/temporalDatabases.tex
First version

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Preliminaries
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section we will introduce some background concepts about possibility theory and temporal databases. Subsection \ref{subsec:poss} is an introduction of possibility theory. The concepts defined by Dubois and Prade~\cite{Dubois1983} about fuzzy numbers and fuzzy intervals are also explained. Temporal databases and their characteristics are introduced in subsection \ref{subsec:temporal}.

\subsection{\label{subsec:poss}Possibilistic variables}
Informally, a theory of uncertainty considers a set of uncertain events and tries to quantify the confidence about the occurence of those events. Consider therefore a set of outcomes $\Omega$ and let each subset of $\Omega$ be called an \emph{event}. Hereby, $\Omega$ is called the \emph{certain event} and can thus be regarded as a point of reference. Similarly, $\emptyset$ is called the \emph{impossible event}. The quantification of confidence in a theory of uncertainty is achieved by means of a \emph{confidence measure}~\cite{Choquet53}.
\begin{definition}
\label{def:conf-measure}
Consider a universe of outcomes $\Omega$. A confidence measure is defined by a function:
\begin{equation}
g:\Pow(\Omega)\rightarrow[0,1]
\end{equation}
that satisfies:
\begin{eqnarray}
g\left(\emptyset\right)&=&0\\
g\left(\Omega\right)&=&1\\
A\subseteq B&\Rightarrow&g(A)\leq g(B)
\end{eqnarray}
and where $\Pow(\Omega)$ represents the power set of $\Omega$.
\end{definition}
From the constraints in Definition \ref{def:conf-measure}, it can be seen that $\emptyset$ and $\Omega$ are indeed respectively the impossible and the certain event. In addition, the confidence that is assigned to events is monotonic. Due to this monotonicity, we have that:
\begin{eqnarray}
\forall A\in\Omega:\forall B\in\Omega&:&g\left(A\cup B\right)\geq\max\left(g\left(A\right),g\left(B\right)\right)\\
\forall A\in\Omega:\forall B\in\Omega&:&g\left(A\cap B\right)\leq\min\left(g\left(A\right),g\left(B\right)\right).
\end{eqnarray}
We can thus derive two extreme measures of confidence and these measures play a crucial role in the theory of possibility~\cite{Dubois88}. Consider a confidence measure $\Pi$ that satisfies the following constraint:
\begin{equation}
\label{EQ:MAXITIVITY}
\forall A\subseteq\Omega:\forall B\subseteq\Omega:\Pi\left(A\cup B\right) = \max\left(\Pi(A),\Pi(B)\right).
\end{equation}
The function $\Pi$ is then called a \emph{possibility measure} and due to \eqref{EQ:MAXITIVITY}, it is said that $\Pi$ is \emph{maxitive}~\cite{Dubois88}. For each event $A\subseteq\Omega$, $\Pi(A)$ expresses the possibility that $A$ occurs. Due to the maxitivity of $\Pi$ and under the constraint that $\Omega$ is finite, we can easily extract information about the possibility of separate outcomes in the form of a \emph{possibility distribution} $\pi$ defined by:
\begin{equation}
\pi:\Omega\rightarrow[0,1]:\pi(u)=\Pi\left(\left\{u\right\}\right).
\end{equation}
Equivalently, we have that:
\begin{equation}
\forall A\subseteq\Omega:\Pi(A)=\sup_{u\in A}\pi (u).
\end{equation}
A possibility distribution is normalized in the sense that:
\begin{equation}
\sup_{u\in\Omega}\pi(u)=1.
\end{equation}
In order to ensure the existence of a possibility distribution when $\Omega$ is infinite, the maxitivity constraint \eqref{EQ:MAXITIVITY} should be replaced by:
\begin{equation}
\Pi\left(\bigcup_{i}A_i\right)=\sup_{i}\Pi\left(A_i\right)
\end{equation}
for any family $\left\{A_i\right\}$ of indexed subsets of $U$~\cite{Nguyen79}.

Let us now consider a confidence measure $\N$ that satisfies:
\begin{equation}
\label{EQ:MINITIVITY}
\forall A\subseteq\Omega:\forall B\subseteq\Omega:\N\left(A\cap B\right) = \min\left(\Pi(A),\Pi(B)\right).
\end{equation}
The function $\N$ is then called a \emph{necessity measure} and due to \eqref{EQ:MINITIVITY}, it is said that $\N$ is \emph{minitive}. Possibility measures and necessity measures are dual in the sense that:
\begin{equation}
\forall A\subseteq\Omega:\Pi(A)=1-\N(\overline{A}).
\end{equation}
The principles of possibility theory are often used in the scope of a \emph{possibilistic variable}, also known as an \emph{ill-known value}.
\begin{definition}
Consider a universe $U$. A possibilistic variable $X$ over $U$ is a variable taking exactly one value in $U$, but for which this value is (partially) unknown. The available knowledge about the value that $X$ takes, is given by a possibility distribution $\pi_X$. For each $u\in U$, $\pi_X(u)$ represents the possibility that $X$ takes the value $u$.
\end{definition}
A specific application of possibility theory~\cite{Prade82,VanSchooten,DeCooman} is obtained when the universe is the set of Boolean values, i.e. $U=\mathbb{B}=\{T,F\}$. If we consider a Boolean proposition, then this proposition is either true ($T$) or false ($F$). Any Boolean proposition $p$ can thus be regarded as a possibilistic variable for which the possibility and necessity that $p=T$ are denoted in short as $\Pos(p)$ and $\Nec(p)$. The following notations are used:
\begin{eqnarray}
\Pos(p)&=&\pi_p(T)\\
\Nec(p)&=&1-\pi_p(F).
\end{eqnarray}
Keeping in mind the cause of uncertainty that is assumed, it is important to understand that a possibilistic variable $X$ is bounded to take one value only, but this value is not known due to incomplete knowledge. When a possibilistic variable is defined over the universe $\Pow(U)$, it is said to be a \emph{possibilistic multi-valued variable}, also known as an \emph{ill-known set}~\cite{Dubois88b}. Note that an ill-known set conceptually differs from a fuzzy set in the sense that an ill-known set is a crisp set, but for some reason (partially) unknown.
\subsection{\label{subsubsec:fuzzy}Fuzzy numbers and fuzzy intervals}
When we shall discuss the case of fuzzy temporal databases, we shall link the concepts introduced in this paper to the theory of fuzzy intervals. Dubois and Prade~\cite{Dubois1983} proposed the following definition of \emph{fuzzy interval}.
\begin{definition}
A fuzzy interval is a fuzzy set $M$ on the set of real numbers $\mathbb{R}$ such that:
\begin{eqnarray}
\forall (u,v)\in\mathbb{R}^2:\forall w \in [u,v]:&\mu_M(w) \geq\min(\mu_M(u),\mu_M(v)) \ &\mbox{ (convexity) } \\
\exists m \in \mathbb{R}:& \mu_M(m)=1 \ &\mbox{ (normalization) }
\end{eqnarray}
\end{definition}
If $m$ is unique, then $M$ is referred to as a \emph{fuzzy number}, instead of a \emph{fuzzy interval}. In other words, if the core of a fuzzy interval is a singleton, it can be seen as a fuzzy number. In their work, Dubois and Prade propose four different functions (two possibility and two necessity functions) to asses the position of a fuzzy number N relative to that of a fuzzy number M taken as a reference.

The most convenient representation for the membership function of a fuzzy number is a triangular function. The membership function $\mu_M$ for the fuzzy set $M$ has also the properties of convexity and normalization. Three values represent a triangular function (see Figure~\ref{fig:triangular}):
\begin{itemize}
\item
$D$ is the singleton value in the core of $\mu_M(x)$.
\item
$D-a$ is the lower value in the support of $\mu_M(x)$. 
\item
$D+b$ is the upper value in the support of $\mu_M(x)$.
\end{itemize}
Note that the values $a$ and $b$ are values in the underlying ordered domain. E.g. $(a,b) \in \mathbb{R}^2$. The notation for a triangular function adopted here is $[D,a,b]$.
\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.4]{graphs/triangular.eps}
  \caption{Triangular function.}
  \label{fig:triangular}
\end{figure}

The most simple representation for the membership function of a fuzzy interval is a trapezoidal function. This membership function $\mu_T$ for the fuzzy interval $T$ is convex and normalized. Four values represent a trapezoidal membership function (figure  \ref{fig:trapezoidal}): $\left[\alpha,\ \beta,\ \gamma,\ \delta\right]$. The membership function is given by:

\begin{align}
\mu_T(x)
\begin{cases}
1 & \mbox{ if } x \in [\beta,\gamma] \\
0 & \mbox{ if } x > \delta \vee x < \alpha \\
\frac{x-\alpha}{\beta - \alpha} & \mbox{ if } x \in [\alpha,\beta[ \\
\frac{\delta -x}{\delta - \gamma} & \mbox{ if } x \in ]\gamma,\delta] \\
\end{cases}
\end{align}

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.4]{graphs/trapezoidalDistribution.eps}
  \caption{Trapezoidal distribution.}
  \label{fig:trapezoidal}
\end{figure}

\subsection{\label{subsec:temporal}Temporal databases}

\subsubsection{\label{subsubsec:timeDomain}Time Domain}
The semantics of time and the different approaches to the time model are analyzed in this section; a more detailed explanation is discussed in \ref{sec:timeindatabasedesign}.

The general time model represents time as a set with an imposed partial order. There are two main models of time: linear and cyclic. In the linear model, time advances from the past to the future in a totally ordered fashion (a relationship of total order is imposed to the set). A cyclic model of time is used for recurrent processes. The present proposal focuses on the linear model.
\begin{itemize}
\item
Discrete model:  The discrete model of time is isomorphic in relation to the natural numbers. According to this model, each natural number corresponds to an indivisible time unit. The said time unit is called a chronon or, in other words, a 'time quantum', 'moment' or 'instant'. A chronon is the smallest time unit that can be represented by the model.
\item
Dense model: This model is isomorphic in relation to the rational or the real numbers. It represents one instant of time in the gap between another two instants.
\item
Continuous model: This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers.
\end{itemize}
Restrictions on time range may be performed on the base of the following two concepts:
\begin{itemize}
\item
Time boundedness: Time can be bounded orthogonally in the past and in the future.
\item
Time distance: As a metric, time has a distance function which presents four properties:
\begin{enumerate}
\item
Distance is non-negative.
\item
The distance between two different elements is not equal to zero.
\item
The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
\item
Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
\end{enumerate}
\end{itemize}
On the other hand, time may be relative or absolute (in other words, anchored or unanchored). Sometimes, what one considers 'absolute time' is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).

\subsubsection{Time in database design}
\label{sec:timeindatabasedesign}
This section defines the model used to represent time in a database. This model allows working with time instants and time intervals.  Even though there is no general agreement as to whether real-time line is continuous or discrete, a discrete model of time may be the most appropriate for a temporal database.

>>>>

%
% Temporal databases
%
This section defines the model used to represent time in a database. This model allows working with time instants and time intervals.  Even though there is no general agreement as to whether real-time line is continuous or discrete, a discrete model of time may be the most appropriate for a temporal database.
